---
title: "Homework 7"
author: "Ryan Corydon and Emma Podlich"
date: "2023-03-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Link to GIBHUB Repository: https://github.com/ryancorydon8/homework7

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
```


```{r}
HAPPY <- readRDS("data/HAPPY.rds")
```

## Part 1 (cleaning)

1. Data cleaning: the values “IAP”, “DK” and “NA” all encode missing values. Replace all of these instances by the value NA.

```{r}
HAPPY[HAPPY == "IAP" | HAPPY == "DK" | HAPPY == "NA"] <- NA
```


2. Check the type of the variable and cast into the right type (factor variable for categorical variables). For age, change “89 OR OLDER” to 89 and assume the variable should be numeric.

3. Bring all levels of factors into a sensible order. For marital you could e.g. order the levels according to average age.

```{r}
str(HAPPY)
```

```{r}
HAPPY$HAPPY <- as.factor(HAPPY$HAPPY)
HAPPY %>% ggplot(aes(x = HAPPY)) + geom_bar()

# This plot shows that order of happiness from greatest to least amount of people for this data set is in the following order: pretty happy, very happy, not too happy, and NA. Most people are pretty happy or very happy in this data set and not as many are unhappy.
```

```{r}
HAPPY$YEAR <- as.numeric(HAPPY$YEAR)
HAPPY %>% ggplot(aes(x = YEAR)) + geom_histogram(binwidth=5)

# This plot shows that there are good variety of years represented in this data set from 1972 to 2016. The greatest amount of data is from the late 1990s-early 2000s but there is still a lot in other years. The least amount of data is from the early 1970s.
```

```{r}
HAPPY$SEX <- as.factor(HAPPY$SEX)
HAPPY %>% ggplot(aes(x = SEX)) + geom_bar()

# This plot shows that there are more females than males in this dataset by about 10000. 
```

```{r}
HAPPY$MARITAL <- as.factor(HAPPY$MARITAL)

HAPPY.M = HAPPY %>% mutate(MARITAL = factor(MARITAL, levels(reorder(MARITAL, AGE, mean, na.rm = TRUE)))) 

HAPPY.M %>% ggplot(aes(x = MARITAL)) + geom_bar()

# This plot shows that order of marital status from greatest to least amount of people for this data set is in the following order: married, never married, divorced, widowed, separated, and NA. Most people in this dataset are either married or never married. Very few people are separated, widowed, or divorced.
```

```{r}
HAPPY$DEGREE <- as.factor(HAPPY$DEGREE)
HAPPY.D = HAPPY %>% mutate(DEGREE = factor(DEGREE, levels=c("LT HIGH SCHOOL", "HIGH SCHOOL", "JUNIOR COLLEGE", "BACHELOR", "GRADUATE SCHOOL")))
HAPPY.D %>% ggplot(aes(x = DEGREE)) + geom_bar()

# This plot shows that order of degree from greatest to least amount of people for this data set is in the following order: high school, lt high school, bachelor, junior college, and NA. Most people in this data set have a high school degree. 
```

```{r}
HAPPY$FINRELA <- as.factor(HAPPY$FINRELA)
HAPPY.F = HAPPY %>% mutate(FINRELA = factor(FINRELA, levels=c("FAR BELOW AVERAGE", "BELOW AVERAGE", "AVERAGE", "ABOVE AVERGAGE", "FAR ABOVE AVERAGE")))
HAPPY.F %>% ggplot(aes(x = FINRELA)) + geom_bar()

# This plot shows that order of finrela from greatest to least amount of people for this data set is in the following order: average, NA, below average, far below average, and far above average. Most people in this data set have an average finrela. Not very many people in this data set have a far below or far above average finrela.
```

```{r}
HAPPY$HEALTH <- as.factor(HAPPY$HEALTH)
HAPPY.H = HAPPY %>% mutate(HEALTH = factor(HEALTH, levels=c("POOR", "FAIR", "GOOD", "EXCELLENT")))
HAPPY.H %>% ggplot(aes(x = HEALTH)) + geom_bar()

# This plot shows that order of health from greatest to least amount of people for this data set is in the following order: good, NA, excellent, fair, and poor. Most people in this data set have good or above health and few have fair and below health.
```

```{r}
HAPPY$WTSSALL <- as.numeric(HAPPY$WTSSALL)
HAPPY %>% ggplot(aes(x = WTSSALL)) + geom_histogram(binwidth = .5)

# This plot shows that the most of the data has wtssall values around 1. There are very few with values above 2.5 and most of the data ranges from values of 0-2.
```

```{r}
HAPPY$PARTYID <- as.factor(HAPPY$PARTYID)
HAPPY.P = HAPPY %>% mutate(PARTYID = factor(PARTYID, levels=c("STRONG DEMOCRAT", "NOT STR DEMOCRAT", "IND,NEAR DEM", "INDEPENDENT", "IND,NEAR REP", "NOT STR REPUBLICAN", "STRONG REPUBLICAN", "OTHER PARTY"))) 
HAPPY.P %>% ggplot(aes(x = PARTYID)) + geom_bar()

# This plot shows that order of partyid from greatest to least amount of people for this data set is in the following order: not str democrat, strong democrat, not str republican, independent, ind near dem, strong republican, ind near rep, other party and NA. There is a good mixture of partyids represented in this data set, however, there are slightly more "left-leaning" people represented in this data set.
```

```{r}
HAPPY$POLVIEWS <- as.factor(HAPPY$POLVIEWS)
HAPPY.PV = HAPPY %>% mutate(POLVIEWS = factor(POLVIEWS, levels=c("EXTREMELY LIBERAL", "LIBERAL", "SLIGHTLY LIBERAL", "MODERATE", "SLIGHTLY CONSERVATIVE", "CONSERVATIVE", "EXTREMELY CONSERVATIVE"))) 
HAPPY.PV %>% ggplot(aes(x = POLVIEWS)) + geom_bar()

# This plot shows that order of political views from greatest to least amount of people for this data set is in the following order: moderate, NA, conservative, slightly liberal, liberal, and extremely liberal. There are a lot of moderate and NA political views in this data set. There are few extreme viewpoints on either side represented in this dataset.
```

```{r}
HAPPY[HAPPY == "89 OR OLDER"] <- "89"
HAPPY$AGE <- as.numeric(HAPPY$AGE)
HAPPY %>% ggplot(aes(x = AGE)) + geom_histogram(binwidth=1)

# This plot shows a decreasing number of people in this data set by age; the most number of people are about 30 and from age 30 the number of people in the data set decreases. From age 18-30 there is an increase in the number of people represented in this data. Overall, there is a lot of younger people in this data set.
```


## Part 2 Individual Exploration

### Ryan

The relationship that I want to investigate is the happiness from the years 2013-2016 of people that are married and not married.

```{r}
HAPPY1 = 
HAPPY %>% filter(YEAR > '2012') %>% filter(MARITAL == 'MARRIED' | MARITAL == 'DIVORCED') 
```

```{r}
 ggplot(HAPPY1, aes(x = YEAR))+
  geom_bar(aes(fill = MARITAL), position = "dodge") + 
  facet_wrap(~HAPPY) 
```

Looking at the data I can see in the through the years 2013 and 2016 I can clearly see that people that are married are much more happy than those that aren't. I can assume that this will be the same for all years as well but wanted to do a specific year range for fun. So if you want to be more happier in life I assume you should try and get married.

### Emma

The relationship that I would like to investigate is how many people in this dataset are happy are based on their different sexes and marital status.

```{r}
ggplot(HAPPY, aes(x = SEX))+
  geom_bar(aes(fill = MARITAL), position = "dodge") + 
  facet_wrap(~HAPPY)
```

According to this plot, the trends in both happiness type and marital status is the same for males and females. There are few males/females in the not too happy category and all martial status' have about the same amount of males/females in this category. There are a lot of males/females in the pretty happy category and most common marital status would be married people are in this category. The least common would be separated in this category. There are a lot of males/females in the very happy category but a little bit less than the previous category (pretty happy). The most common marital status would also be married and the least common would also be separated in this category. Overall, the majority of people in this data set are pretty happy or very happy and married. There appears to be slightly more females than males in this data set but the number is fairly close. Males/females also have very similar happiness levels based on their martial status.