---
title: "Homework 7"
author: "Ryan Corydon and Emma Podlich"
date: "2023-03-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
```


```{r}
#data(happy, package="productplots")
#head(happy)
```
```{r}
HAPPY <- readRDS("data/HAPPY.rds")
```



1. Data cleaning: the values “IAP”, “DK” and “NA” all encode missing values. Replace all of these instances by the value NA.

```{r}
HAPPY[HAPPY == "IAP" | HAPPY == "DK" | HAPPY == "NA"] <- NA
```


2. Check the type of the variable and cast into the right type (factor variable for categorical variables). For age, change “89 OR OLDER” to 89 and assume the variable should be numeric.

```{r}
HAPPY[HAPPY=="89 OR OLDER"] <- "89"
HAPPY$AGE <- as.numeric(HAPPY$AGE)
```



3. Bring all levels of factors into a sensible order. For marital you could e.g. order the levels according to average age.





Part 2 Individual Exploration

Ryan

The relationship that I want to investigate is the happiness from the years 2008-2016 of people that are married and not married.

```{r}
HAPPY1 = 
HAPPY %>% filter(YEAR > '2012') %>% filter(MARITAL == 'MARRIED' | MARITAL == 'DIVORCED') 
```

```{r}
 ggplot(HAPPY1, aes(x = YEAR))+
  geom_bar(aes(fill = MARITAL), position = "dodge") + 
  facet_wrap(~HAPPY) 
```
Looking at the data I can see in the through the years 2013 and 2016 I can clearly see that people that are married are much more happy than those that aren't. 

Emma